Viscous Transport in Eroding Porous Media

### 2019-10-09 docs/transport/erosion_transport.tex
starting to move to JFM format

\documentclass[preprint,10pt]{elsarticle}

\input{preambles.tex}
%\usepackage{showkeys}
\begin{document}

@@

% OTHER TITLE POSSIBILITIES
% ???

\author[SH]{Shang-Huan Chiu}
\author[Nick]{M.~N.~J.~Moore}
\author[Bryan]{Bryan D.~Quaife}

\address[SH]{Department of Scientific Computing, Florida State
University, Tallahassee, FL, 32306.}
\address[Nick]{Department of Mathematics and Geophysical Fluid Dynamics
Institute, Florida State University, Tallahassee, FL, 32306.}
\address[Bryan]{Department of Scientific Computing and Geophysical Fluid
Dynamics Institute, Florida State University, Tallahassee, FL, 32306.}

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through specific two-dimensional porous geometries---those that have
  become channelized by the process of fluid-mechanical erosion.  We
  consider passive tracers in the eroded geometries and analyze their to
  statistical moments to investigate the effect of erosion on the
  tortuosity and dispersion rates. We compare the computed
  porosity-tortuosity relationship with models available in the
  literature. Finally, we analyze the pore size distribution of an
  eroding geometry. The simulations are performed by combining a
  high-fidelity boundary integral equation solver for the Stokes
  equations, a second-order stable time stepping method to simulate
  erosion, a new numerical method to stably and accurately compute the
  eroded geometries and particle trajectories.

@@

\begin{keyword}
  Porous media \sep Stokes flow \sep Erosion \sep Transport \sep
  Tortuosity \sep Anomalous dispersion \sep Pore size distributions \sep
  Quadrature
\end{keyword}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998} and velocity scales
can be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}.
Moreover, for a single porous geometry, the pore sizes and velocities
can range over several orders of magnitude.  Numerical methods that
resolve this range of scales offer the ability to: (i) characterize
dispersion~\cite{saf1959}, (ii) quantify
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and
(iii) develop meaningful constitutive relationships that link the
microscopic and macroscopic realms~\cite{mil-chr-imh-mcb-ped1998}.
Examples of coarse-grained models for porous media flow include
pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, pore and flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is further complicated when boundaries dynamically
evolve in response to the fluid flow. This coupling between the geometry
and the flow occurs, for example, in applications involving
melting~\cite{bec-vis1998, rycroft2016asymmetric, jambon2018singular,
favier2019rayleigh, morrow2019moving},
dissolution~\cite{kan-zha-che-he2002, mac2015shape, moo2017,
wykes2018self}, deposition~\cite{joh-eli1995, hewett2018modelling},
biofilm growth~\cite{tan-val-wer2015}, and crack
formation~\cite{cho2019crack}. We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, and
industrial applications~\cite{berhanu2012shape, hewett2017evolution,
lachaussee2018competitive, lopez2018cfd, allen2019sde, amin2019role}.
When a porous geometry erodes, certain qualitative characteristics are
unveiled that affect transport through the geometry.  For example, an
eroded geometry may contain channels of high porosity, which, though few
in number and modest in volume fraction, transmit a large portion of the
flux \cite{qua-moo2018}. This arrangement results in velocities that
vary over several orders of magnitude~\cite{all-hea-lab-rei2002}.
Moreover, channelization creates heterogeneous and anisotropic medium
properties, which affect the transport of tracers such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterization of
tracer transport through the resulting eroded geometries. Our modeling
efforts build on previous work~\cite{ris-moo-chi-she-zha2012,
moore2013self, moo2017}, in particular recent numerical methods
developed to simulate erosion in the Stokes-flow
regime~\cite{qua-moo2018}. We, however, make key improvements to the
numerical methods from~\cite{qua-moo2018} to enable simulations of more
realistic, dense suspensions of eroding grains
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  Then, to
characterize transport through these configurations, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

\begin{figure}[htp]
\begin{center}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b1}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b80}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b160}\\
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b240}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b320}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b400}
\caption{\label{fig:Eroding50vel} Six evenly spaced snapshots of 50
eroding bodies.  The background velocity is a Hagen-Poiseuille flow with
a varying flow strength that corresponds to a constant pressure drop
across the channel. The color is the magnitude of the velocity in a
logarithmic scale. The flow velocity varies over several orders of
magnitude, the grains form irregular shapes with large aspect ratios,
and the geometry becomes channelized and anisotropic.}
\end{center}
\end{figure}

Owing to the scales present in groundwater flow~\cite{bea1972}, we model
the hydrodynamics with the two-dimensional incompressible Stokes
equations.  Meanwhile, individual grains erode at a rate proportional to
the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  Since the
vorticity and the shear stress are equivalent on solid boundaries, we
also compute the fluid vorticity for visualization purposes
(Figure~\ref{fig:Eroding50vort}).  A variety of numerical methods for
the fluid solver are available including lattice-gas or lattice
Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, simulating
erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration method is unchanged from
previous work~\cite{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be triggered by changes in
sign of the shear stress, and we use a stable second-order Runge-Kutta
method applied to the {\thL} coordinates~\cite{hou-low-she1994} of the
eroding grains.  In this work, we introduce a new quadrature method to
resolve dense suspensions.  The accuracy of the
trapezoid rule, which was used in previous work, is adequate for
bodies that are sufficiently separated~\cite{tre-wei2014}.  However, it
loses accuracy whenever grains are in close contact.  One of the
earliest works to address quadrature methods for nearly-singular
integrands was developed by Baker and Shelley~\cite{bak-she1986}, and in
recent years, many other near-singular integration schemes have
followed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\cite{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate the shear stress and
the fluid vorticity.

\begin{figure}[htp]
\begin{center}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b1}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b80}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b160}\\
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b240}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b320}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b400}
\caption{\label{fig:Eroding50vort} The same six snapshots as
Figure~\ref{fig:Eroding50vel}.  The color is the vorticity of the fluid.
Since the rate of erosion is equivalent to the magnitude of the
vorticity, erosion is fastest in the yellow and blue regions and slowest
in the green regions.}
\end{center}
\end{figure}

To characterize transport through the resulting configurations, particle
trajectories must be computed. Depending on the application, microscale
transport can be modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories to be governed by pure advection
(ie.~no diffusion), so the particle trajectories are identical to the
streamlines. We compute trajectories $\ss(t)$ that are initialized at
$\ss_0$ by solving the advection equation
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term in~\eqref{eqn:tracers}, we use a fourth-order explicit Runge-Kutta
time stepping method.  In addition, the Barycentric quadrature rule is
used to accurately compute trajectories that are close to an eroding
grain. We note that an alternative approach to compute trajectories is
to form a distribution of velocities $\uu$ and use a continuous time
random walk~\cite{mon-wei1965, den-cor-sch-ber2004, leb-den-car2008,
ber-cor-den-sch2006}.

Once the streamlines are computed, we characterize transport by
analyzing three different metrics: the tortuosity, the anomalous
dispersion, and the pore size distribution.  The local tortuosity of a
streamline that connects the inlet to the outlet is defined as the
streamline's length normalized by the linear inlet-to-outlet distance.
In porous media, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008} or even
2~\cite{dud-koz-mat2011}, depending on several factors such as the
porosity. The tortuosity of a geometry is defined by averaging the local
tortuosity over all streamlines initialized at the inlet, and the
geometry's tortuosity characterizes average particle
motions~\cite{hak-com-den2019}.  To characterize spreading, the fluid
dispersion is computed by calculating the variance of the streamline
lengths. In porous media, this spreading is often
super-dispersive~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. Since anomalous dispersion results from
streamlines spending time in both the high and low velocity
regimes~\cite{ber-sch2001}, it is crucial to accurately resolve
streamlines near grain boundaries, as achieved in this work.  Finally,
we construct the pore-size distribution throughout the erosion process.
These distributions are required to quantify velocity
distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019},
connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first contribution is extending the Barycentric quadrature rule of
Barnett et al.~\cite{bar-wu-vee2015} to accurately compute velocity
gradients associated with Stokes flow. This new quadrature rule permits
accurate and stable simulations of erosion, and, moreover, enables
visualization of the vorticity field throughout.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\cite{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this work does not limit the proximity
of  grains. Compared to our previous work, we are able to simulate much
denser suspensions with a comparable number of discretization points and
without upsampling.  While we do investigate transport in this work, we
only consider passive particles that are advected by the flow.  To model
other physical processes, such as contaminant transport, we plan to
simulate a finite Peclet number advection-diffusion equation in eroded
geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the erosion model that is
described in more detail in previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry and transport. Section~\ref{sec:method} describes the numerical
methods, with special attention paid to the new quadrature method for
computing the shear stress and the vorticity.  Section~\ref{sec:results}
presents numerical examples for a variety of dense packings of bodies.
Finally, concluding remarks are made in Section~\ref{sec:conclusions}.

>>>>

%\documentclass[preprint,10pt]{elsarticle}
\documentclass{jfm}

\input{preambles.tex}
%\usepackage{showkeys}

\usepackage{graphicx}
\usepackage{epstopdf, epsfig}

\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\shorttitle{Viscous Transport in Eroding Porous Media}
\shortauthor{Chiu, Moore and Quaife}

@@

\author{Shang-Huan Chiu\aff{1}, M.~N.~J.~Moore\aff{2}, and Bryan
Quaife\aff{3}\corresp{\email{bquaife@fsu.edu}}}

%  Alan N. Other\aff{1}
%  \corresp{\email{jfm@damtp.cam.ac.uk}},
%  H. - C. Smith\aff{1}
% \and J. Q.  Public\aff{2}}

\affiliation{
\aff{1}Department of Scientific Computing, Florida State University,
Florida State University, Tallahassee, FL 32306, USA
\aff{2}Department of Mathematics and Geophysical Fluid Dynamics
Institute, Florida State University, Tallahassee, FL 32306, USA
\aff{3}Department of Scientific Computing and Geophysical Fluid Dynamics
Institute, Florida State University, Tallahassee, FL 32306, USA
}

\begin{document}

%\title{Viscous Transport in Eroding Porous Media}
%
%\author[SH]{Shang-Huan Chiu}
%\author[Nick]{M.~N.~J.~Moore}
%\author[Bryan]{Bryan D.~Quaife}
%
%\address[SH]{Department of Scientific Computing, Florida State
%University, Tallahassee, FL, 32306.}
%\address[Nick]{Department of Mathematics and Geophysical Fluid Dynamics
%Institute, Florida State University, Tallahassee, FL, 32306.}
%\address[Bryan]{Department of Scientific Computing and Geophysical Fluid
%Dynamics Institute, Florida State University, Tallahassee, FL, 32306.}

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through specific two-dimensional porous geometries---those that are
  formed by fluid-mechanical erosion.  We investigate the tortuosity and
  dispersion by analyzing the first two statistical moments of tracer
  trajectories. For most geometries, the tortuosity decreases as erosion
  increases the porosity; however, interestingly, it is possible for
  erosion to increase the tortuosity. The simulated porosity-tortuosity
  relationships are compared with models available in the literature.
  Unsurprisingly, the asymptotic dispersion rate is also affected by the
  erosion process, but it is also affected by the number of and
  distribution of the eroding bodies.  Finally, we analyze the pore size
  distribution of an eroding geometry. The simulations are performed by
  combining a high-fidelity boundary integral equation solver for the
  fluid equations, a second-order stable time stepping method to
  simulate erosion, and new numerical methods to stably and accurately
  resolve nearly-touching eroded bodies and particle trajectories near
  the eroding bodies.

@@

%\begin{keyword}
%  Porous media \sep Stokes flow \sep Erosion \sep Transport \sep
%  Tortuosity \sep Anomalous dispersion \sep Pore size distributions \sep
%  Quadrature
%\end{keyword}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\citep{mil-chr-imh-mcb-ped1998} and velocity scales
can be as small as $10^{-1}\mathrm{m/day}$~\citep{kut-scr-dav-ham1995}.
Moreover, for a single porous geometry, the pore sizes and velocities
can range over several orders of magnitude.  Numerical methods that
resolve this range of scales offer the ability to: (i) characterize
dispersion~\citep{saf1959}, (ii) quantify
mixing~\citep{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and
(iii) develop meaningful constitutive relationships that link the
microscopic and macroscopic realms~\citep{mil-chr-imh-mcb-ped1998}.
Examples of coarse-grained models for porous media flow include
pressure-saturation-permeability
relationships~\citep{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\citep{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\citep{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\citep{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\citep{ica-den2019, bel-sal-rin1992}, pore and flow
distributions~\citep{ali-par-wei-bre2017}, geometry
connectivity~\citep{knu-car2005, wes-blo-gra2001}, anomalous
dispersion~\citep{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is further complicated when boundaries dynamically
evolve in response to the fluid flow. This coupling between the geometry
and the flow occurs, for example, in applications involving
melting~\citep{bec-vis1998, rycroft2016asymmetric, jambon2018singular,
favier2019rayleigh, morrow2019moving},
dissolution~\citep{kan-zha-che-he2002, mac2015shape, moo2017,
wykes2018self}, deposition~\citep{joh-eli1995, hewett2018modelling},
biofilm growth~\citep{tan-val-wer2015}, and crack
formation~\citep{cho2019crack}. We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, and
industrial applications~\citep{berhanu2012shape, hewett2017evolution,
lachaussee2018competitive, lopez2018cfd, allen2019sde, amin2019role}.
When a porous geometry erodes, certain qualitative characteristics are
unveiled that affect transport through the geometry.  For example, an
eroded geometry may contain channels of high porosity, which, though few
in number and modest in volume fraction, transmit a large portion of the
flux \citep{qua-moo2018}. This arrangement results in velocities that
vary over several orders of magnitude~\citep{all-hea-lab-rei2002}.
Moreover, channelization creates heterogeneous and anisotropic medium
properties, which affect the transport of tracers such as
contaminants~\citep{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\citep{nil-sto1990, ree-sto1995}.

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterization of
tracer transport through the resulting eroded geometries. Our modeling
efforts build on previous work~\citep{ris-moo-chi-she-zha2012,
moore2013self, moo2017}, in particular recent numerical methods
developed to simulate erosion in the Stokes-flow
regime~\citep{qua-moo2018}. We, however, make key improvements to the
numerical methods from~\citep{qua-moo2018} to enable simulations of more
realistic, dense suspensions of eroding grains
(figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  Then, to
characterize transport through these configurations, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

\begin{figure}
\begin{center}
\includegraphics[height = 0.3 \textwidth]{./figs/vel_log_50b1}
\includegraphics[height = 0.3 \textwidth]{./figs/vel_log_50b80}
\includegraphics[height = 0.3 \textwidth]{./figs/vel_log_50b160}\\
\includegraphics[height = 0.3 \textwidth]{./figs/vel_log_50b240}
\includegraphics[height = 0.3 \textwidth]{./figs/vel_log_50b320}
\includegraphics[height = 0.3 \textwidth]{./figs/vel_log_50b400}
\caption{\label{fig:Eroding50vel} Six evenly spaced snapshots of 50
eroding bodies.  The background velocity is a Hagen-Poiseuille flow with
a varying flow strength that corresponds to a constant pressure drop
across the channel. The color is the magnitude of the velocity in a
logarithmic scale. The flow velocity varies over several orders of
magnitude, the grains form irregular shapes with large aspect ratios,
and the geometry becomes channelized and anisotropic.}
\end{center}
\end{figure}

Owing to the scales present in groundwater flow~\citep{bea1972}, we model
the hydrodynamics with the two-dimensional incompressible Stokes
equations.  Meanwhile, individual grains erode at a rate proportional to
the hydrodynamic shear stress~\citep{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  Since the
vorticity and the shear stress are equivalent on solid boundaries, we
also compute the fluid vorticity for visualization purposes
(figure~\ref{fig:Eroding50vort}).  A variety of numerical methods for
the fluid solver are available including lattice-gas or lattice
Boltzmann~\citep{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\citep{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\citep{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\citep{rot1988}.  However, simulating
erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration method is unchanged from
previous work~\citep{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be triggered by changes in
sign of the shear stress, and we use a stable second-order Runge-Kutta
method applied to the {\thL} coordinates~\citep{hou-low-she1994} of the
eroding grains.  In this work, we introduce a new quadrature method to
resolve dense suspensions.  The accuracy of the
trapezoid rule, which was used in previous work, is adequate for
bodies that are sufficiently separated~\citep{tre-wei2014}.  However, it
loses accuracy whenever grains are in close contact.  One of the
earliest works to address quadrature methods for nearly-singular
integrands was developed by Baker and Shelley~\citep{bak-she1986}, and in
recent years, many other near-singular integration schemes have
followed~\citep{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\citep{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate the shear stress and
the fluid vorticity.

\begin{figure}
\begin{center}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b1}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b80}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b160}\\
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b240}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b320}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b400}
\caption{\label{fig:Eroding50vort} The same six snapshots as
figure~\ref{fig:Eroding50vel}.  The color is the vorticity of the fluid.
Since the rate of erosion is equivalent to the magnitude of the
vorticity, erosion is fastest in the yellow and blue regions and slowest
in the green regions.}
\end{center}
\end{figure}

To characterize transport through the resulting configurations, particle
trajectories must be computed. Depending on the application, microscale
transport can be modelled as pure advection~\citep{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\citep{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\citep{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories to be governed by pure advection
(ie.~no diffusion), so the particle trajectories are identical to the
streamlines. We compute trajectories $\ss(t)$ that are initialized at
$\ss_0$ by solving the advection equation
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term in~\eqref{eqn:tracers}, we use a fourth-order explicit Runge-Kutta
time stepping method.  In addition, the Barycentric quadrature rule is
used to accurately compute trajectories that are close to an eroding
grain. We note that an alternative approach to compute trajectories is
to form a distribution of velocities $\uu$ and use a continuous time
random walk~\citep{mon-wei1965, den-cor-sch-ber2004, leb-den-car2008,
ber-cor-den-sch2006}.

Once the streamlines are computed, we characterize transport by
analyzing three different metrics: the tortuosity, the anomalous
dispersion, and the pore size distribution.  The local tortuosity of a
streamline that connects the inlet to the outlet is defined as the
streamline's length normalized by the linear inlet-to-outlet distance.
In porous media, the local tortuosity can be greater than
1.5~\citep{kop-kat-tim1996, mat-kha-koz2008} or even
2~\citep{dud-koz-mat2011}, depending on several factors such as the
porosity. The tortuosity of a geometry is defined by averaging the local
tortuosity over all streamlines initialized at the inlet, and the
geometry's tortuosity characterizes average particle
motions~\citep{hak-com-den2019}.  To characterize spreading, the fluid
dispersion is computed by calculating the variance of the streamline
lengths. In porous media, this spreading is often
super-dispersive~\citep{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. Since anomalous dispersion results from
streamlines spending time in both the high and low velocity
regimes~\citep{ber-sch2001}, it is crucial to accurately resolve
streamlines near grain boundaries, as achieved in this work.  Finally,
we construct the pore-size distribution throughout the erosion process.
These distributions are required to quantify velocity
distributions~\citep{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\citep{sie-ili-pri-riv-gua2019},
connectivity~\citep{knu-car2005, wes-blo-gra2001}, and to develop network
models~\citep{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first contribution is extending the Barycentric quadrature rule of
Barnett et al.~\citep{bar-wu-vee2015} to accurately compute velocity
gradients associated with Stokes flow. This new quadrature rule permits
accurate and stable simulations of erosion, and, moreover, enables
visualization of the vorticity field throughout.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\citep{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this work does not limit the proximity
of  grains. Compared to our previous work, we are able to simulate much
denser suspensions with a comparable number of discretization points and
without upsampling.  While we do investigate transport in this work, we
only consider passive particles that are advected by the flow.  To model
other physical processes, such as contaminant transport, we plan to
simulate a finite Peclet number advection-diffusion equation in eroded
geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the erosion model that is
described in more detail in previous work~\citep{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry and transport. Section~\ref{sec:method} describes the numerical
methods, with special attention paid to the new quadrature method for
computing the shear stress and the vorticity.  Section~\ref{sec:results}
presents numerical examples for a variety of dense packings of bodies.
Finally, concluding remarks are made in Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\citep{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of the boundary of $[-3,3] \times [-1,1]$.  We only
place eroding bodies in $[-1,1] \times [-1,1]$ to create a buffer region
for the flow profile imposed at the inlet to transition to the more
complex flow intervening between the bodies. On each eroding body, $\nn$
is the normal vector pointing into the body, and $\ss$ is the unit
tangent vector pointing in the counterclockwise direction.  The normal
vector on $\Gamma$ is the outward unit normal.  Neglecting inertial
forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$. We simulate erosion by alternating
between solving the fluid equations and advancing the erodible
boundaries. The strength of $\UU$ is adjusted at each time step to
achieve a constant pressure drop across the channel.
